### 2018-06-16 paper/5-implementations.tex
Work on S5

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Make build system
make :: Ord k => Build Applicative (MakeInfo k) k v
make = topological modTimeRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build information: current time and file modification times
type Time       = Integer
type MakeInfo k = (Time, Map k Time)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on file modification times
modTimeRebuilder :: Ord k => Rebuilder Applicative (MakeInfo k) k v
modTimeRebuilder key value task = Task $ \fetch -> do
    (now, modTime) <- get
    let dirty = case Map.lookup key modTime of
            Nothing -> True
            time -> any (\d -> Map.lookup d modTime > time) (dependencies task)
    if not dirty
    then return value
    else do
        put (now + 1, Map.insert key now modTime)
        run task fetch
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A topological task scheduler
topological :: Ord k => Rebuilder Applicative i k v -> Build Applicative i k v
topological rebuilder tasks target = execState $ forM_ order $ \key -> case tasks key of
    Nothing   -> return ()
    Just task -> do
        value <- gets (getValue key)
        let newTask :: Task (MonadState i) k v
            newTask = rebuilder key value task
            newFetch :: k -> StateT i (State (Store i k v)) v
            newFetch = lift . gets . getValue
        info <- gets getInfo
        (newValue, newInfo) <- runStateT (run newTask newFetch) info
        modify $ putInfo newInfo . putValue key newValue
  where
    deps k = case tasks k of { Nothing -> [@@]; Just task -> dependencies task }
    order  = case topSort (reachable deps target) of
        Nothing -> error "Cannot build tasks with cyclic dependencies"
        Just xs -> xs
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard graph algorithms (implementation omitted)
reachable :: Ord k => (k -> [k]) -> k -> Graph k
topSort   :: Ord k => Graph k -> Maybe [k]
\end{minted}
\vspace{-2mm}
\caption{An implementation of \Make using our framework.}\label{fig-make-implementation}
\vspace{-4mm}
\end{figure}

@@

We capture persistent build information that \Make stores by the pair
\hs{(@@now,}~\hs{modTime)} comprising the \emph{current time} \hs{now} and the
dictionary \hs{modTime}~\hs{::}~\hs{Map}~\hs{k}~\hs{Time} that stores \emph{file
modification times}. The rebuilder receives three arguments: a \hs{key}, its
current \hs{value}, and the \hs{task} of type
\hs{Task}~\hs{Applicative}~\hs{k}~\hs{v} that can be used to rebuild the
\hs{key} if necessary. The rebuilder returns a modified monadic task of type
\hs{Task}~\hs{(@@MonadState}~\hs{(@@MakeInfo}~\hs{k))}~\hs{k}~\hs{v} that has
access to the build information and uses it to decide whether to return the
current \hs{value} as is or rebuild the \hs{key} if it is \hs{dirty}, i.e.
if it is older than any of its dependencies\footnote{Since the \hs{task} is
applicative we can statically find its dependencies using the function
\hs{dependencies} defined in~\S\ref{sec-deps}.} or has never been built before
(in which case \hs{modTime} has no entry for it). Note that in our model we
update the current time and file modification time whenever the \hs{key} is
rebuilt, which in reality is taken care of by the file system.

\Make's scheduler processes keys in a linear \hs{order} based on a
\hs{topological} sort of the statically known dependency graph
(see~\S\ref{sec-parallelism} for parallel \Make). Our definition in
Fig.~\ref{fig-make-implementation} is polymorphic with respect to the type of
build information \hs{i} and is therefore compatible with any applicative
\hs{rebuilder}. The scheduler calls the supplied \hs{rebuilder} on every
\hs{key} in the \hs{order}, and runs the obtained \hs{newTask} to compute the
\hs{newValue} and updated build information \hs{newInfo}, which are subsequently
written to the store. The computation runs in the
\hs{State}~\hs{(@@Store}~\hs{i}~\hs{k}~\hs{v)} monad, similar to the \hs{busy}
build system (\S\ref{sec-general-build}), but we also need to use an additional
\hs{StateT}~\hs{i} wrapper to run the \hs{newTask} since it only has access to
the \hs{i} part of the \hs{Store}~\hs{i}~\hs{k}~\hs{v}.

@@

We define \Excel, with its reordering dependency strategy, in  Fig.~\ref{fig-excel-implementation}.
\Excel's persistently stored information is a triple: (i) the dirty bit
function \hs{k}~\hs{->}~\hs{Bool}, (ii) an approximation of key dependencies
\hs{k}~\hs{->}~\hs{ApproximateDependencies}~\hs{k} that \Excel uses to handle
dynamic dependencies, and (iii) the calculation chain \hs{[@@k]} recorded in the
previous build (\S\ref{sec-background-excel}).

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Approximate task dependencies: a map of known upper bounds of dependencies
type ApproximateDependencies k = Map k [k]
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build information: dirty keys, approximate dependencies, and calculation chain
type CalcChain k = [k]
type ExcelInfo k = ((Set k, ApproximateDependencies k), CalcChain k)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Excel build system
excel :: Ord k => Build Monad (ExcelInfo k) k v
excel = restarting approximateRebuilder
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A restarting task scheduler
restarting :: Ord k => Rebuilder Monad i k v -> Build Monad (i, Chain k) k v
restarting rebuilder tasks key = execState $ do
    chain    <- gets (snd . getInfo)
    newChain <- go Set.empty $ chain ++ [key | key `notElem` chain]
    modify . mapInfo $ \(i, _) -> (i, newChain)
  where
    go :: Set k -> Chain k -> State (Store (i, Chain k) k v) (Chain k)
    go _    []     = return []
    go done (k:ks) = case tasks k of
        Nothing -> (k :) <$> go (Set.insert k done) ks
        Just task -> do
            store <- get
            let value = getValue k store
                newTask :: Task (MonadState i) k (Either k v)
                newTask = try $ rebuilder k value task -- `try' is defined below
                newFetch :: k -> State i (Either k v)
                newFetch k | k `Set.member` done = return $ Right (getValue k store)
                           | otherwise           = return $ Left k
            case runState (run newTask newFetch) (fst $ getInfo store) of
                (Left dep, _) -> go done $ [ dep | dep `notElem` ks ] ++ ks ++ [k]
                (Right newValue, newInfo) -> do
                    modify $ putInfo (newInfo, []) . updateValue k value newValue
                    (k :) <$> go (Set.insert k done) ks
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Convert a total task into a task accepting a partial fetch callback
try :: Task (MonadState i) k v -> Task (MonadState i) k (Either e v)
try task = Task $ \fetch -> runExceptT $ run task (ExceptT . fetch)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on approximate dependencies
approximateRebuilder :: Ord k => Rebuilder Monad (Set k, ApproximateDependencies k) k v
approximateRebuilder key value task = Task $ \fetch -> do
    (isDirty, deps) <- get
    let dirty = key `Set.member` isDirty || case Map.lookup key deps of
                                                Nothing -> True
                                                Just ks -> any (`Set.member` isDirty) ks
    if not dirty
    then return value
    else do
        put (Set.insert key isDirty, deps)
        run task fetch
\end{minted}
\vspace{-2mm}
\caption{An implementation of \Excel using our framework.}\label{fig-excel-implementation}
\vspace{-2mm}
\end{figure}

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Shake build system
shake :: (Ord k, Hashable v) => Build Monad (VT k v) k v
shake = suspending vtRebuilder
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A suspending task scheduler
suspending :: Ord k => Rebuilder Monad i k v -> Build Monad i k v
suspending rebuilder tasks key store = fst $ execState (fetch key) (store, Set.empty)
  where
    fetch :: k -> State (Store i k v, Set k) v
    fetch key = case tasks key of
        Nothing -> gets (getValue key . fst)
        Just task -> do
            done <- gets snd
            when (key `Set.notMember` done) $ do
                value <- gets (getValue key . fst)
                let newTask :: Task (MonadState i) k v
                    newTask = rebuilder key value task
                    newFetch :: k -> StateT i (State (Store i k v, Set k)) v
                    newFetch k = do v <- lift $ fetch k
                                    put =<< lift (gets (getInfo . fst))
                                    return v
                info <- gets (getInfo . fst)
                (newValue, newInfo) <- runStateT (run newTask newFetch) info
                modify $ \(s, _) ->
                    (putInfo newInfo $ putValue key newValue s, Set.insert key done)
            gets (getValue key . fst)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Rebuilder based on verifying traces
vtRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (VT k v) k v
vtRebuilder key value task = Task $ \fetch -> do
    vt <- get
    dirty <- not <$> verifyVT key value (fmap hash . fetch) vt
    if not dirty
    then return value
    else do
        (newValue, deps) <- trackM task fetch
        put =<< recordVT key newValue deps (fmap hash . fetch) =<< get
        return newValue
\end{minted}
\vspace{-3mm}
\caption{An implementation of \Shake using our framework.}\label{fig-shake-implementation}
\vspace{1mm}
\end{figure}

>>>>

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Make build system
make :: Ord k => Build Applicative (MakeInfo k) k v
make = topological modTimeRebuilder
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build information: current time and file modification times
type Time       = Integer
type MakeInfo k = (Time, Map k Time)
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on file modification times
modTimeRebuilder :: Ord k => Rebuilder Applicative (MakeInfo k) k v
modTimeRebuilder key value task = Task $ \fetch -> do
    (now, modTime) <- get
    let dirty = case Map.lookup key modTime of
            Nothing -> True
            time -> any (\d -> Map.lookup d modTime > time) (dependencies task)
    if not dirty then return value else do
        put (now + 1, Map.insert key now modTime)
        run task fetch
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- A topological task scheduler
topological :: Ord k => Rebuilder Applicative i k v -> Build Applicative i k v
topological rebuilder tasks target = execState $ forM_ order $ \key -> case tasks key of
    Nothing   -> return ()
    Just task -> do
        value <- gets (getValue key)
        let newTask :: Task (MonadState i) k v
            newTask = rebuilder key value task
            fetch :: k -> StateT i (State (Store i k v)) v
            fetch = lift . gets . getValue
        info <- gets getInfo
        (newValue, newInfo) <- runStateT (run newTask fetch) info
        modify $ putInfo newInfo . putValue key newValue
  where
    deps k = case tasks k of { Nothing -> [@@]; Just task -> dependencies task }
    order  = case topSort (reachable deps target) of
        Nothing -> error "Cannot build tasks with cyclic dependencies"
        Just xs -> xs
\end{minted}
\vspace{0mm}
\begin{minted}[fontsize=\small]{haskell}
-- Standard graph algorithms (implementation omitted)
reachable :: Ord k => (k -> [k]) -> k -> Graph k
topSort   :: Ord k => Graph k -> Maybe [k]
\end{minted}
\vspace{-2mm}
\caption{An implementation of \Make using our framework.}\label{fig-make-implementation}
\vspace{-4mm}
\end{figure}

@@

We capture persistent build information that \Make stores by the pair
\hs{(@@now,}~\hs{modTime)} comprising the \emph{current time} \hs{now} and the
map \hs{modTime}~\hs{::}~\hs{Map}~\hs{k}~\hs{Time} that stores \emph{file
modification times}. The rebuilder receives three arguments: a \hs{key}, its
current \hs{value}, and the \hs{task} of type
\hs{Task}~\hs{Applicative}~\hs{k}~\hs{v} that can be used to rebuild the
\hs{key} if necessary. The rebuilder returns a modified monadic task of type
\hs{Task}~\hs{(@@MonadState}~\hs{(@@MakeInfo}~\hs{k))}~\hs{k}~\hs{v} that has
access to the build information and uses it to decide whether to return the
current \hs{value} as is or rebuild the \hs{key} if it is \hs{dirty}, i.e.
if it is older than any of its dependencies\footnote{Since the \hs{task} is
applicative we can statically find its dependencies using the function
\hs{dependencies} defined in~\S\ref{sec-deps}.} or has never been built before
(in which case \hs{modTime} has no entry for it). Note that in our model we
update the current time and file modification time whenever the \hs{key} is
rebuilt, which in reality is taken care of by the file system.

\Make's scheduler processes keys in a linear \hs{order} based on a
\hs{topological} sort of the statically known dependency graph
(see~\S\ref{sec-parallelism} for parallel \Make). Our definition in
Fig.~\ref{fig-make-implementation} is polymorphic with respect to the type of
build information \hs{i} and is therefore compatible with any applicative
\hs{rebuilder}. The scheduler calls the supplied \hs{rebuilder} on every
\hs{key} in the \hs{order}, and runs the obtained \hs{newTask} to compute the
\hs{newValue} and updated build information \hs{newInfo}, which are subsequently
written to the store. The computation runs in the
\hs{State}~\hs{(@@Store}~\hs{i}~\hs{k}~\hs{v)} monad, similar to the \hs{busy}
build system (\S\ref{sec-general-build}), but we need to use an additional
\hs{StateT}~\hs{i} wrapper to run the \hs{newTask} since it only has access to
the \hs{i} part of the \hs{Store}~\hs{i}~\hs{k}~\hs{v}. The \hs{newTask} can
find values of the \hs{key}'s dependencies via the \hs{fetch} callback that
reads the store by executing \hs{getValue} in the inner state monad.

@@

Fig.~\ref{fig-excel-implementation} shows an implementation of \Excel in our
framework as a combination of the \hs{restarting} scheduler and an
\hs{approximateRebuilder}.

\Excel's persistently stored information is a triple: (i) the set of dirty keys
\hs{Set}~\hs{k}, (ii) a map storing known upper bounds of dependencies
\hs{Map}~\hs{k}~\hs{[@@k]} that \Excel uses to handle dynamic dependencies, and
(iii) the calculation chain \hs{[@@k]} recorded in the previous build
(\S\ref{sec-background-excel}).

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Excel build system
excel :: Ord k => Build Monad (ExcelInfo k) k v
excel = restarting approximateRebuilder
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Approximate task dependencies: a map of known upper bounds of dependencies
type ApproximateDependencies k = Map k [k]
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A task rebuilder based on the set of dirty keys and approximate dependencies
approximateRebuilder :: Ord k => Rebuilder Monad (Set k, ApproximateDependencies k) k v
approximateRebuilder key value task = Task $ \fetch -> do
    (isDirty, deps) <- get
    let dirty = key `Set.member` isDirty || case Map.lookup key deps of
                                                Nothing -> True
                                                Just ks -> any (`Set.member` isDirty) ks
    if not dirty then return value else do
        put (Set.insert key isDirty, deps)
        run task fetch
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Build information: dirty keys, approximate dependencies, and calculation chain
type Chain k = [k]
type ExcelInfo k = ((Set k, ApproximateDependencies k), Chain k)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A restarting task scheduler
restarting :: Ord k => Rebuilder Monad i k v -> Build Monad (i, Chain k) k v
restarting rebuilder tasks key = execState $ do
    chain    <- gets (snd . getInfo)
    newChain <- go Set.empty $ chain ++ [key | key `notElem` chain]
    modify . mapInfo $ \(i, _) -> (i, newChain)
  where
    go :: Set k -> Chain k -> State (Store (i, Chain k) k v) (Chain k)
    go _    []     = return []
    go done (k:ks) = case tasks k of
        Nothing -> (k :) <$> go (Set.insert k done) ks
        Just task -> do
            store <- get
            let value = getValue k store
                newTask :: Task (MonadState i) k (Either k v)
                newTask = try $ rebuilder k value task -- `try' is defined below
                fetch :: k -> State i (Either k v)
                fetch k | k `Set.member` done = return $ Right (getValue k store)
                        | otherwise           = return $ Left k
            case runState (run newTask fetch) (fst $ getInfo store) of
                (Left dep, _) -> go done $ [ dep | dep `notElem` ks ] ++ ks ++ [k]
                (Right newValue, newInfo) -> do
                    modify $ putInfo (newInfo, []) . updateValue k value newValue
                    (k :) <$> go (Set.insert k done) ks
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Convert a total task into a task accepting a partial fetch callback
try :: Task (MonadState i) k v -> Task (MonadState i) k (Either e v)
try task = Task $ \fetch -> runExceptT $ run task (ExceptT . fetch)
\end{minted}
\vspace{-2mm}
\caption{An implementation of \Excel using our framework.}\label{fig-excel-implementation}
\vspace{-2mm}
\end{figure}

@@

\begin{figure}
\begin{minted}[fontsize=\small]{haskell}
-- Shake build system
shake :: (Ord k, Hashable v) => Build Monad (VT k v) k v
shake = suspending vtRebuilder
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- A suspending task scheduler
suspending :: Ord k => Rebuilder Monad i k v -> Build Monad i k v
suspending rebuilder tasks key store = fst $ execState (build key) (store, Set.empty)
  where
    build :: k -> State (Store i k v, Set k) v
    build key = case tasks key of
        Nothing -> gets (getValue key . fst)
        Just task -> do
            done <- gets snd
            when (key `Set.notMember` done) $ do
                value <- gets (getValue key . fst)
                let newTask :: Task (MonadState i) k v
                    newTask = rebuilder key value task
                    fetch :: k -> StateT i (State (Store i k v, Set k)) v
                    fetch k = do v <- lift (build k)
                                 put =<< lift (gets (getInfo . fst))
                                 return v
                info <- gets (getInfo . fst)
                (newValue, newInfo) <- runStateT (run newTask fetch) info
                modify $ \(s, _) ->
                    (putInfo newInfo $ putValue key newValue s, Set.insert key done)
            gets (getValue key . fst)
\end{minted}
\vspace{1mm}
\begin{minted}[fontsize=\small]{haskell}
-- Rebuilder based on verifying traces
vtRebuilder :: (Eq k, Hashable v) => Rebuilder Monad (VT k v) k v
vtRebuilder key value task = Task $ \fetch -> do
    vt <- get
    dirty <- not <$> verifyVT key value (fmap hash . fetch) vt
    if not dirty
    then return value
    else do
        (newValue, deps) <- trackM task fetch
        put =<< recordVT key newValue deps (fmap hash . fetch) =<< get
        return newValue
\end{minted}
\vspace{-3mm}
\caption{An implementation of \Shake using our framework.}\label{fig-shake-implementation}
\vspace{1mm}
\end{figure}
